RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-12-29 manuscripts/manuscript.Rmd
Tightening up read/write section.

## Reading phylogenetic trees

We begin by reading in an example NeXML file provided with the
package, containing two small phylogenetic trees.[^2]

@@

The method `nexml_read()` reads NeXML files, either from a local
file, or from a remote location via its URL, and returns an object of
type `nexml`. This object type is a representation of the NeXML
document contents, and therefore holds phylogenetic trees as well as 
character or trait matrices, and all metadata. This is similar in
design to the classes in the `phylobase` package, which contain 
trees and traits but hold only limited metadata [@phylobase], but
different from R object types such as `ape::phylo` (the `phylo` class
in the `ape` package [@Paradis_2004]), which represents trees alone.

@@

```{r}
phy <- get_trees_list(nex)
class(phy)
```

As an example for leveraging the wealth of resources in R
interoperating with the `ape::phylo` type, we use `ape`'s standard
plotting function to plot the first tree in the list of trees:

```{r eval=FALSE}
plot(phy[[1]]) 
```

@@

Because the example file supplied with the package contains two
phylogenies in a single `trees` block, in this case the return type of both
phylogeny accessor methods is the same:

```{r fig.show="hide"}
phy <- get_trees(nex)
plot(phy[[1]]) 
```

As mentioned above, the `nexml_read()` method allows specifying the
NeXML document to read not only as the name of a local file, but also
as the URL of a document on the web. If the location parameter is
recognized as a URL, the method will automatically download the
document to the local working directory and read it from there. This
gives convenient and rapid access to phylogenetic data published in
NeXML format on the web, which includes the content of the
phylogenetic data repository TreeBASE [@Piel_2002]. For
example, the following plots a tree in TreeBASE:

@@

## Reading character data

NeXML documents can contain character or trait data in the form of
matrices. A data matrix can be of molecular (for molecular sequence
alignments), discrete (for most morphological character data), or
continuous type (for many trait data). Character data, especially
molecular and morphological matrices, are often used for phylogenetic
tree reconstruction, but they are also a common input for the many
comparative phylogenetic analysis methods available in R. The method
`get_characters()` obtains character data from a `nexml` object, and
returns them as a standard `data.frame` R object with columns as
characters and rows as taxa. The following example, which uses a file
that comes with `RNeXML`, demonstrates this:

@@

Observe that `RNeXML` has automatically merged the different NeXML character
blocks, (representing continuous and discrete characters respectively) 
into a single data frame with appropriate column classes. 
To allow strict validation of data types, the NeXML schema requires
continuous and discrete character data to be separated into different
"characters" blocks. Since the `data.frame` data structure in R has no
such constraints, the `get_characters()` method will combine such
blocks as separate columns into a single `data.frame` object, provided
they correspond to the same taxa. Otherwise, a list of `data.frame`s
is returned, with list elements corresponding to characters
blocks. Similar to the distinction between `get_trees()` and
`get_trees_list()`, there is also a method `get_characters_list()`,
which always returns a list of `data.frame`s, one for each
character block.

@@

Aside from providing an easy-to-use interface to data already in
NeXML, `RNeXML` also aims to facilitate broader adoption of the
standard for sharing phylogenetic data, and thus to make it
straightforward to generate them. The method `nexml_write()`, as the
name suggests, generates a NeXML file from its input parameters. In
its simplest invocation, the method writes a tree to a file:

@@

In addition to trees, the `nexml_write()` method also to specify
character data as another parameter. The following example uses data
from the comparative phylogenetics R package `geiger` [@Harmon_2008].

>>>>

## Reading and writing phylogenies and character data

Conceptually, a NeXML document has the following components: (1)
phylogeny data, (2) character or trait data in matrix form, (3)
operational taxonomic units (OTUs), and (4) metadata. When reading and
writing NeXML, `RNeXML` aims to map these to and from, respectively,
their most natural representations in R. These are the `phylo` and
`multiPhylo` objects from the `ape` package [@Paradis_2004] for
phylogenies, and R's native `data.frame` list structure for character
matrices.

## Reading phylogenies and character data

The method `nexml_read()` reads NeXML files, either from a local
file, or from a remote location via its URL, and returns an object of
type `nexml` [^2]:

@@

The `nexml` object type is a representation of the NeXML
document contents, and therefore holds phylogenetic trees as well as 
character or trait matrices, and all metadata. This is similar in
design to the classes in the `phylobase` package, which contain 
trees and traits but hold only limited metadata [@phylobase], but
different from R object types such as `ape::phylo` (the `phylo` class
in the `ape` package [@Paradis_2004]), which represents trees alone.

@@

```{r}
phy <- get_trees_list(nex)
```

@@

If the location parameter with which the `nexml_read()` method is
invoked is recognized as a URL, the method will automatically download the
document to the local working directory and read it from there. This
gives convenient and rapid access to phylogenetic data published in
NeXML format on the web, such as the content of the
phylogenetic data repository TreeBASE [@Piel_2002]. For
example, the following plots a tree in TreeBASE:

@@

The method `get_characters()` obtains character data matrics from a
`nexml` object, and returns them as a standard `data.frame` R object
with columns as characters and rows as taxa:

@@

A NeXML data matrix can be of molecular (for molecular sequence
alignments), discrete (for most morphological character data), or
continuous type (for many trait data).  Similarly as for trees, data
matrices can also be arranged in blocks; in fact, matrices of
different data types are required to be separated into different
"blocks" to allow strict validation of data typs. Since the
`data.frame` data structure in R has no such constraints, the
`get_characters()` method will combine such blocks as separate columns
into a single `data.frame` object, provided they correspond to the
same taxa. Otherwise, a list of `data.frame`s is returned, with list
elements corresponding to characters blocks. Similar to the methods
for obtaining trees, there is also a method `get_characters_list()`,
which always returns a list of `data.frame`s, one for each character
block.

@@

The method `nexml_write()` generates a NeXML file from its input parameters. In
its simplest invocation, the method writes a tree to a file:

@@

In addition to trees, the `nexml_write()` method also allows to specify
character data as another parameter. The following example uses data
from the comparative phylogenetics R package `geiger` [@Harmon_2008].
